Inferring the existence and direction of causal associations in the face of measurement error

### 2017-02-18 docs/manuscript.Rmd
updated

The analysis proceeds by simulating variables that have known causal effects and valid instruments, and then applying measurement error in different configurations. We then assess the performance of the CIT test and MR analysis for a) inference of a causal relationship and b) inferring the correct direction of causality. Included in the MR analysis is a simple extension that is used to make inference of the causal direction when it is unknown which of the correlated variables the instrument is directly acting upon. All analysis was performed using the R programming language [@RCoreTeam2015] and code is made available at [github url to go here]() and implemented in the MR-Base (http://wwww.mrbase.org) platform [@Hemani2016].

@@

### Steiger sensitivity analysis

The Steiger test for inferring if $x \rightarrow y$ is based on evaluating $\rho_{gx} > \rho_{gy}$. However, $\rho_{gx}$ and $\rho_{gy}$ are underestimated if $x$ and $y$ are measured imprecisely. If, for example, $x$ has lower measurement precision than $y$ then we might empirically obtain $\rho_{g,x_o} > \rho_{g,y_o}$. As we show in the appendix, the maximum measurement error of $x_o$ is $\rho_{g,x_o}$ and the minimum is 0 (same applies to $y_o$). It is possible to simulate what the inferred causal direction would be for all values within these bounds. To evaluate how reliable, $R$, the Steiger test is to potential measurement imprecision in $x$ and $y$ we integrate over the simulated $R$ values for the possible measurement error values which are compatible with the direction of causality obtained empirically from the Steiger test. We also do the same for the measurement error values which are not compatible. The ratio of these quantities is an indication of how much of the measurement error space supports the inferred causal direction. A ratio $R=1$ indicates that the Steiger test is highly sensitive to measurement error, with equal weight of the measurement error parameter space supporting both directions of causality. Full details are provided in Appendix 2 and `r cfig("steiger_sensitivity_plot")`a.

@@

@@

### Using MR to infer the direction of causality

If we do not know whether the SNP $g$ has a direct influence on $x$ or $y$ then further efforts are required. If $x$ causes $y$ and $g$ is a valid instrument for $x$ then we expect that $\rho_{g,x}$ to be greater than $\rho_{g,y}$ because the association of $g$ with $y$ is the proportion of the true causal association between $x$ and $y$ scaled by the association between $g$ and $x$. We can use the Steiger test of correlated correlations to formally test the difference in magnitude of $\rho_{g,x}$ and $\rho_{g,y}$, where the variable with the largest correlation on $g$ is deemed to be the variable which is imposing the causal effect.

This is a simple method to orient the direction of causality in an MR analysis when the underlying biology of the SNP is not fully understood. But it can be shown that in the presence of measurement imprecision, $d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}$ (Appendix 2), where $d$ is a metric of the degree to which the Steiger test is liable to provide the wrong estimate (*i.e.* if $d>0$ then the Steiger test is likely to be correct about the causal direction). Thus it is important to note that under certain conditions when measurement imprecision is present the Steiger test could make erroneous inference about the causal direction because the value of $d$ is not restricted to being greater than 0. However `r cfig("d_relationship_figure")` shows that when there is no measurement error in $x$, the Steiger test is unlikely to infer the wrong direction of causality even if there is measurement error in $y$. It also shows that in most cases where $x$ is measured with error, especially when the causal effect between $x$ and $y$ is not very large, the sensitivity of the Steiger test to measurement error is relatively low.

@@

The Steiger test is used to infer if the variant $g$ has a direct influence on $x$ or $y$ when it is known that it associates with both, but the direction of causality between $x$ and $y$ is unknown. Assuming the causal direction is $x \to y$, two stage MR is formulated using the following regression models:

@@

In the presence of measurement error in $x$ and $y$, however, the Steiger test will instead be assessing the inequality $\rho_{g, x_o} > \rho_{g, y_o}$, which can be simplified:

@@

In order to assess how reliable the Steiger test is in the presence of measurement imprecision, we can evaluate the range of potential values of measurement error in $x$ and $y$ over which the empirical result from the Steiger test would return the wrong value.

>>>>

@@

### Causal direction sensitivity analysis

The Steiger test for inferring if $x \rightarrow y$ is based on evaluating $\rho_{gx} > \rho_{gy}$. However, $\rho_{gx}$ and $\rho_{gy}$ are underestimated if $x$ and $y$ are measured imprecisely. If, for example, $x$ has lower measurement precision than $y$ then we might empirically obtain $\rho_{g,x_o} > \rho_{g,y_o}$. As we show in the appendix, the maximum measurement error of $x_o$ is $\rho_{g,x_o}$ and the minimum is 0 (same applies to $y_o$). It is possible to simulate what the inferred causal direction would be for all values within these bounds. To evaluate how reliable, $R$, the inference of the causal direction is to potential measurement imprecision in $x$ and $y$ we integrate over the simulated $R$ values for the possible measurement error values which are compatible with the direction of causality obtained empirically from the Steiger test. We also do the same for the measurement error values which are not compatible. The ratio of these quantities is an indication of how much of the measurement error space supports the inferred causal direction. A ratio $R=1$ indicates that the Steiger test is highly sensitive to measurement error, with equal weight of the measurement error parameter space supporting both directions of causality. Full details are provided in Appendix 2 and `r cfig("steiger_sensitivity_plot")`a.

@@

All analysis was performed using the R programming language [@RCoreTeam2015] and code is made available at [github url to go here]() and implemented in the MR-Base (http://wwww.mrbase.org) platform [@Hemani2016].

@@

### Using MR Steiger to infer the direction of causality

If we do not know whether the SNP $g$ has a primary influence on $x$ or $y$ then CIT can attempt to infer the causal direction. Here we introduce the MR Steiger approach to also orient the direction of causality in an MR analysis when the underlying biology of the SNP is not fully understood.

We show that in the presence of measurement imprecision, $d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}$ (Appendix 2), where $d$ is a metric of the degree to which the Steiger test is liable to provide the wrong estimate (*i.e.* if $d>0$ then the Steiger test is likely to be correct about the causal direction). Thus it is important to note that under certain conditions when measurement imprecision is present the Steiger test could make erroneous inference about the causal direction because the value of $d$ is not restricted to being greater than 0. However `r cfig("d_relationship_figure")` shows that when there is no measurement error in $x$, the Steiger test is unlikely to infer the wrong direction of causality even if there is measurement error in $y$. It also shows that in most cases where $x$ is measured with error, especially when the causal effect between $x$ and $y$ is not very large, the sensitivity of the Steiger test to measurement error is relatively low.

@@

Assuming that either $x \rightarrow y$ or $y \rightarrow x$, the causal direction can be inferred by evaluating which of $\rho_{g, x}$ and $\rho_{g, y}$ is larger in magnitude. The Steiger test is a hypothesis test that provides a p-value for observing the difference in these correlations under the null hypothesis that they are equal.

Assuming the causal direction is $x \to y$, two stage MR is formulated using the following regression models:

@@

In the presence of measurement error in $x$ and $y$, however, the empirical inference of the causal direction will instead be based on evaluating $\rho_{g, x_o} > \rho_{g, y_o}$, which can be simplified:

@@

In order to assess how reliable the inference of the causal direction is in the presence of measurement imprecision, we can evaluate the range of potential values of measurement error in $x$ and $y$ over which the empirical difference in $\rho_{g, x_o}$ and $\rho_{g, y_o}$ would return the wrong causal direction.

### 2017-02-18 docs/manuscript.tex
updated

\setlength{\droptitle}{-2em}
  \title{Inferring the existence and direction of causal associations in the face
of measurement error}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{15 February 2017}

@@

where \(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o\)

MR analysis estimates the causal influence of \(x\) on \(y\) by using
the instrument as a proxy for \(x\), such that

@@

where \(\beta_{MR} \neq 0\).

Measurement error of an exposure can be modeled as some transformation
of the true value that leads to the observed value, \(x_o = f(x)\). For
example, we can define
\(f(x) = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx}\), where
\(\alpha_{mx}\) and \(\beta_{mx}\) influence the error in the
measurement of \(x\) by altering its scale, and \(\epsilon_{mx}\)
represents the imprecision (or noise) in the measurement of \(x\). Here
the true value of the exposure is partially explained by the genetic
instrument, \(g\), such that

where \(\beta_g\) is the effect of the SNP on the exposure, and
\(\epsilon_g\) is the normally distributed residual value of \(x\); and
the outcome is partially explained by the exposure

where \(\beta_x\) is the true effect of the exposure on the outcome. In
the causal inference test (CIT), an omnibus p-value is generated from
four hypothesis tests: 1) \(g\) is associated with \(x\); 2) \(g\) is
associated with \(y\); 3) \(x\) is associated with \(y\); and 4) \(g\)
is independent of \(y|x\). The 4th condition employs mediation for
causal inference, and can be expressed as \(cov(g, y - \hat{y}) = 0\),
where \(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o\). When measurement
error in scale and imprecision is introduced, such that \(y_o\) is the
measured value of \(y\), it can be shown using basic covariance
properties (Appendix 1) that

When selecting an instrument \(g\) that has a direct influence on \(x\)
where \(x\) is causally related to \(y\), one can reason that the
influence of \(g\) on \(y\) is the proportion of the effect of \(x\) on
\(y\) that is explained by the effect of \(g\) on \(x\). Hence the
effect estimate \(\beta_{MR}\) of \(x\) on \(y\) is estimated as

\[
\begin{aligned}
\beta_{MR} = \frac{\beta_{y_o}}{\beta_g} & = \frac{cov(y_o, g)}{cov(x_o, g)} \\
                                         & = \frac{\beta_{my} \beta_x cov(x, g)} {\beta_{mx} cov(x, g)} \\
                                         & = \frac{\beta_{my}} {\beta_{mx}} \beta_x
\end{aligned}
\]

Thus we obtain an estimate of the effect of \(x_o\) on \(y\) that is
scaled by the measurement error in \(x\) and \(y\), but is unrelated to
measurement imprecision. Measurement error and imprecision are likely to
reduce the power of MR inferring a causal association (increased false
negative rate), but importantly the evidence for causal association,
based on rejecting the null hypothesis that \(\beta_{MR} = 0\), is not
inflated when the null hypothesis is true.

@@

The analysis proceeds by simulating variables that have known causal
effects and valid instruments, and then applying measurement error in
different configurations. We then assess the performance of the CIT test
and MR analysis for a) inference of a causal relationship and b)
inferring the correct direction of causality. Included in the MR
analysis is a simple extension that is used to make inference of the
causal direction when it is unknown which of the correlated variables
the instrument is directly acting upon. All analysis was performed using
the R programming language\textsuperscript{35} and code is made
available at \href{}{github url to go here} and implemented in the
MR-Base (\url{http://wwww.mrbase.org}) platform\textsuperscript{26}.

@@

\subsubsection{Steiger sensitivity
analysis}\label{steiger-sensitivity-analysis}

The Steiger test for inferring if \(x \rightarrow y\) is based on
evaluating \(\rho_{gx} > \rho_{gy}\). However, \(\rho_{gx}\) and
\(\rho_{gy}\) are underestimated if \(x\) and \(y\) are measured
imprecisely. If, for example, \(x\) has lower measurement precision than
\(y\) then we might empirically obtain \(\rho_{g,x_o} > \rho_{g,y_o}\).
As we show in the appendix, the maximum measurement error of \(x_o\) is
\(\rho_{g,x_o}\) and the minimum is 0 (same applies to \(y_o\)). It is
possible to simulate what the inferred causal direction would be for all
values within these bounds. To evaluate how reliable, \(R\), the Steiger
test is to potential measurement imprecision in \(x\) and \(y\) we
integrate over the simulated \(R\) values for the possible measurement
error values which are compatible with the direction of causality
obtained empirically from the Steiger test. We also do the same for the
measurement error values which are not compatible. The ratio of these
quantities is an indication of how much of the measurement error space
supports the inferred causal direction. A ratio \(R=1\) indicates that
the Steiger test is highly sensitive to measurement error, with equal
weight of the measurement error parameter space supporting both
directions of causality. Full details are provided in Appendix 2 and
Figure 3a.

@@

@@

\subsubsection{Using MR to infer the direction of
causality}\label{using-mr-to-infer-the-direction-of-causality}

If we do not know whether the SNP \(g\) has a direct influence on \(x\)
or \(y\) then further efforts are required. If \(x\) causes \(y\) and
\(g\) is a valid instrument for \(x\) then we expect that \(\rho_{g,x}\)
to be greater than \(\rho_{g,y}\) because the association of \(g\) with
\(y\) is the proportion of the true causal association between \(x\) and
\(y\) scaled by the association between \(g\) and \(x\). We can use the
Steiger test of correlated correlations to formally test the difference
in magnitude of \(\rho_{g,x}\) and \(\rho_{g,y}\), where the variable
with the largest correlation on \(g\) is deemed to be the variable which
is imposing the causal effect.

This is a simple method to orient the direction of causality in an MR
analysis when the underlying biology of the SNP is not fully understood.
But it can be shown that in the presence of measurement imprecision,
\(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\) (Appendix 2), where \(d\)
is a metric of the degree to which the Steiger test is liable to provide
the wrong estimate (\emph{i.e.} if \(d>0\) then the Steiger test is
likely to be correct about the causal direction). Thus it is important
to note that under certain conditions when measurement imprecision is
present the Steiger test could make erroneous inference about the causal
direction because the value of \(d\) is not restricted to being greater
than 0. However Figure 7 shows that when there is no measurement error
in \(x\), the Steiger test is unlikely to infer the wrong direction of
causality even if there is measurement error in \(y\). It also shows
that in most cases where \(x\) is measured with error, especially when
the causal effect between \(x\) and \(y\) is not very large, the
sensitivity of the Steiger test to measurement error is relatively low.

@@

We also observed that for associations where methylation caused gene
expression the causal effect was more likely to be negative than for the
associations where gene expression caused methylation (OR = 0.61 (95\%
CI 0.36 - 1.03), Figure 6c), suggesting that reducing gene expression
levels at a controlling CpG typically leads to increased gene expression
levels, consistent with expectation\textsuperscript{41}.

@@

Researchers are often confronted with the problem of making causal
inferences using a statistical framework on observational data. In the
epidemiological literature issues of measurement error in mediation
analysis is relatively well explored\textsuperscript{42}. Our analysis
extends this to related methods such as CIT that are widely used in
predominantly 'omic data. These methods are indeed liable to the same
issues as standard mediation based analysis, and specifically we show
that as measurement error in the exposure variable increases, CIT is
likely to have reduced statistical power, and liable to infer the wrong
direction of causality. We also demonstrate that, though unintuitive,
increasing sample size does not resolve the issue, rather it leads to
more extreme p-values for the model that predicts the wrong direction of
causality.

@@

We demonstrate this new method by evaluating the causal relationships of
458 known associations between DNA methylation and gene expression
levels using summary level data. The causal direction is heavily
influenced by how much measurement error is present in the different
assaying platforms. For example, if DNA methylation measures typically
have higher measuremet error than gene expression measures then our
analysis suggests that DNA methylation levels would be more often the
causal factor in the association. Indeed, previous studies which have
evaluated measurement error in these platforms do support this
position\textsuperscript{43,44}, though making strong conclusions for
this analysis is difficult because measurement error is likely to be
study specific.

In our simulations we focused on the simple case of a single instrument
in a single sample setting with a view to making a fair comparison
between MR and the various mediation-based methods available. However,
MR strategies exist for other scenarios, for example if there are
multiple instruments for the hypothesised exposure then the problems
described here may be circumvented\textsuperscript{16}. Likewise, if
there is at least one instrument for each trait then bi-directional MR
can offer solutions to inferring the causal
direction\textsuperscript{45}. In this work we assumed that pleiotropy
(the influence of the instrument on the outcome through a mechanism
other than the exposure) was not present. Recent method developments in
MR\textsuperscript{24,25} have focused on accounting for the issues that
horizontal pleiotropy can introduce, but how they perform in the
presence of measurement error remains to be explored. An important
advantage that MR confers over most mediation based analysis is that it
can be performed in two samples, which can considerably improve power
and expand the scope of analysis. However, whether there is a
substantive difference in two sample MR versus one sample MR in how
measurement error has an effect is not yet fully understood. We have
also assumed no measurement error in the genetic instrument, which is
not unreasonable given the strict QC protocols that ensure high quality
genotype data is available to most studies. We have restricted the scope
to only exploring non-differential measurement error and avoided the
complications incurred if measurement error in the exposure and outcome
is correlated, however our analysis goes beyond many previous
explorations of measurement error by assessing the impacts of both
imprecision (noise) and linear transformations of the true variable on
causal inference.

Mediation based network approaches, that go beyond analyses of two
variables, are very well established\textsuperscript{33} and have a
number of extensions that make them valuable tools, including for
example network construction. But because they are predicated on the
basic underlying principles of mediation they are liable to suffer from
the same issues of measurement error. Recent advances in MR methodology,
for example applying MR to genetical genomics\textsuperscript{46},
multivariate MR\textsuperscript{47} and mediation through
MR\textsuperscript{48--50} may offer more robust alternatives for these
more complicated problems.

@@

The Steiger test is used to infer if the variant \(g\) has a direct
influence on \(x\) or \(y\) when it is known that it associates with
both, but the direction of causality between \(x\) and \(y\) is unknown.
Assuming the causal direction is \(x \to y\), two stage MR is formulated
using the following regression models:

@@

In the presence of measurement error in \(x\) and \(y\), however, the
Steiger test will instead be assessing the inequality
\(\rho_{g, x_o} > \rho_{g, y_o}\), which can be simplified:

@@

In order to assess how reliable the Steiger test is in the presence of
measurement imprecision, we can evaluate the range of potential values
of measurement error in \(x\) and \(y\) over which the empirical result
from the Steiger test would return the wrong value.

@@

\hypertarget{ref-Millstein2016}{}
18. Millstein, J. cit: Causal Inference Test. R package version 1.9.
(2016). at
\textless{}\url{http://cran.r-project.org/package=cit}\textgreater{}

@@

\hypertarget{ref-RCoreTeam2015}{}
35. R Core Team. R: A Language and Environment for Statistical
Computing. (2015). at
\textless{}\url{https://www.r-project.org/}\textgreater{}

@@

\hypertarget{ref-Revelle2015}{}
39. Revelle, W. psych: Procedures for Psychological, Psychometric, and
Personality Research. (2015). at
\textless{}\url{http://cran.r-project.org/package=psych}\textgreater{}

@@

\hypertarget{ref-Bird2002}{}
41. Bird, A. DNA methylation patterns and epigenetic memory. \emph{Genes
\& development} \textbf{16,} 6--21 (2002).

\hypertarget{ref-Cole2014}{}
42. Cole, D. A. \& Preacher, K. J. Manifest Variable Path Analysis:
Potentially Serious and Misleading Consequences Due to Uncorrected
Measurement Error. \emph{Psychological Methods} \textbf{19,} 300--315
(2014).

\hypertarget{ref-Bose2014}{}
43. Bose, M. \emph{et al.} Evaluation of microarray-based DNA
methylation measurement using technical replicates: the Atherosclerosis
Risk In Communities (ARIC) Study. \emph{BMC Bioinformatics} \textbf{15,}
312 (2014).

\hypertarget{ref-Bryant2011}{}
44. Bryant, P. A. \emph{et al.} Technical Variability Is Greater than
Biological Variability in a Microarray Experiment but Both Are
Outweighed by Changes Induced by Stimulation. \emph{PLoS ONE}
\textbf{6,} e19556 (2011).

\hypertarget{ref-Richmond2014}{}
45. Richmond, R. C. \emph{et al.} Assessing Causality in the Association
between Child Adiposity and Physical Activity Levels: A Mendelian
Randomization Analysis. \emph{PLoS Medicine} \textbf{11,} e1001618
(2014).

\hypertarget{ref-Relton2012}{}
46. Relton, C. L. \& Davey Smith, G. Two-step epigenetic Mendelian
randomization: a strategy for establishing the causal role of epigenetic
processes in pathways to disease. \emph{International journal of
epidemiology} \textbf{41,} 161--76 (2012).

\hypertarget{ref-Burgess2014a}{}
47. Burgess, S., Freitag, D. F., Khan, H., Gorman, D. N. \& Thompson, S.
G. Using multivariable Mendelian randomization to disentangle the causal
effects of lipid fractions. \emph{PloS one} \textbf{9,} e108891 (2014).

\hypertarget{ref-Varbo2015}{}
48. Varbo, A. \emph{et al.} Remnant cholesterol, low-density lipoprotein
cholesterol, and blood pressure as mediators from obesity to ischemic
heart disease. \emph{Circulation research} \textbf{116,} 665--73 (2015).

\hypertarget{ref-Burgess2015}{}
49. Burgess, S., Daniel, R. M., Butterworth, A. S. \& Thompson, S. G.
Network Mendelian randomization: using genetic variants as instrumental
variables to investigate mediation in causal pathways.
\emph{International journal of epidemiology} \textbf{44,} 484--95
(2015).

\hypertarget{ref-Richmond2016}{}
50. Richmond, R. C., Hemani, G., Tilling, K., Davey Smith, G. \& Relton,
C. L. Challenges and novel approaches for investigating molecular
mediation. \emph{Human molecular genetics} \textbf{25,} R149--R156
(2016).

>>>>

\setlength{\droptitle}{-2em}
  \title{Inferring the existence and direction of causal associations in the face
of measurement error}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{18 February 2017}

@@

where \(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o\). MR analysis
estimates the causal influence of \(x\) on \(y\) by using the instrument
as a proxy for \(x\), such that

@@

where \(\beta_{MR} \neq 0\) denotes the existence of causality, and the
\(\beta_{MR}\) is an estimate of the causal effect.

Measurement error of an exposure can be modeled as a transformation of
the true value that leads to the observed value, \(x_o = f(x)\). For
example, following Pierce and VanderWeele\textsuperscript{35} we can
define

\[
f(x) = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx}
\]

where \(\alpha_{mx}\) and \(\beta_{mx}\) influence the error in the
measurement of \(x\) by altering its scale, and \(\epsilon_{mx}\)
represents the imprecision (or noise) in the measurement of \(x\). The
same model of measurement error can be applied to the outcome variable
\(y\). In this study we assume there is no measurement error in the SNP.

@@

@@

\subsubsection{Causal direction sensitivity
analysis}\label{causal-direction-sensitivity-analysis}

The Steiger test for inferring if \(x \rightarrow y\) is based on
evaluating \(\rho_{gx} > \rho_{gy}\). However, \(\rho_{gx}\) and
\(\rho_{gy}\) are underestimated if \(x\) and \(y\) are measured
imprecisely. If, for example, \(x\) has lower measurement precision than
\(y\) then we might empirically obtain \(\rho_{g,x_o} > \rho_{g,y_o}\).
As we show in the appendix, the maximum measurement error of \(x_o\) is
\(\rho_{g,x_o}\) and the minimum is 0 (same applies to \(y_o\)). It is
possible to simulate what the inferred causal direction would be for all
values within these bounds. To evaluate how reliable, \(R\), the
inference of the causal direction is to potential measurement
imprecision in \(x\) and \(y\) we integrate over the simulated \(R\)
values for the possible measurement error values which are compatible
with the direction of causality obtained empirically from the Steiger
test. We also do the same for the measurement error values which are not
compatible. The ratio of these quantities is an indication of how much
of the measurement error space supports the inferred causal direction. A
ratio \(R=1\) indicates that the Steiger test is highly sensitive to
measurement error, with equal weight of the measurement error parameter
space supporting both directions of causality. Full details are provided
in Appendix 2 and Figure 3a.

@@

All analysis was performed using the R programming
language\textsuperscript{41} and code is made available at
\href{}{github url to go here} and implemented in the MR-Base
(\url{http://wwww.mrbase.org}) platform\textsuperscript{26}.

@@

\subsubsection{Mediation-based causal inference under measurement
error}\label{mediation-based-causal-inference-under-measurement-error}

In the causal inference test (CIT), an omnibus p-value is generated from
four hypothesis tests: 1) \(g\) is associated with \(x\); 2) \(g\) is
associated with \(y\); 3) \(x\) is associated with \(y\); and 4) \(g\)
is independent of \(y|x\). The 4th condition employs mediation for
causal inference, and can be expressed as \(cov(g, y - \hat{y}) = 0\),
where \(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o\). When measurement
error in scale and imprecision is introduced, such that \(y_o\) is the
measured value of \(y\), it can be shown using basic covariance
properties (Appendix 1) that

\subsubsection{Using MR Steiger to infer the direction of
causality}\label{using-mr-steiger-to-infer-the-direction-of-causality}

If we do not know whether the SNP \(g\) has a primary influence on \(x\)
or \(y\) then CIT can attempt to infer the causal direction. Here we
introduce the MR Steiger approach to also orient the direction of
causality in an MR analysis when the underlying biology of the SNP is
not fully understood.

We show that in the presence of measurement imprecision,
\(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\) (Appendix 2), where \(d\)
is a metric of the degree to which the Steiger test is liable to provide
the wrong estimate (\emph{i.e.} if \(d>0\) then the Steiger test is
likely to be correct about the causal direction). Thus it is important
to note that under certain conditions when measurement imprecision is
present the Steiger test could make erroneous inference about the causal
direction because the value of \(d\) is not restricted to being greater
than 0. However Figure 7 shows that when there is no measurement error
in \(x\), the Steiger test is unlikely to infer the wrong direction of
causality even if there is measurement error in \(y\). It also shows
that in most cases where \(x\) is measured with error, especially when
the causal effect between \(x\) and \(y\) is not very large, the
sensitivity of the Steiger test to measurement error is relatively low.

@@

We also observed that for associations where methylation caused gene
expression the causal effect was more likely to be negative than for the
associations where gene expression caused methylation (OR = 0.61 (95\%
CI 0.36 - 1.03), Figure 6c), suggesting that reducing gene expression
levels at a controlling CpG typically leads to increased gene expression
levels, consistent with expectation\textsuperscript{42}.

@@

Researchers are often confronted with the problem of making causal
inferences using a statistical framework on observational data. In the
epidemiological literature issues of measurement error in mediation
analysis is relatively well explored\textsuperscript{43}. Our analysis
extends this to related methods such as CIT that are widely used in
predominantly 'omic data. These methods are indeed liable to the same
issues as standard mediation based analysis, and specifically we show
that as measurement error in the exposure variable increases, CIT is
likely to have reduced statistical power, and liable to infer the wrong
direction of causality. We also demonstrate that, though unintuitive,
increasing sample size does not resolve the issue, rather it leads to
more extreme p-values for the model that predicts the wrong direction of
causality.

@@

We demonstrate this new method by evaluating the causal relationships of
458 known associations between DNA methylation and gene expression
levels using summary level data. The causal direction is heavily
influenced by how much measurement error is present in the different
assaying platforms. For example, if DNA methylation measures typically
have higher measuremet error than gene expression measures then our
analysis suggests that DNA methylation levels would be more often the
causal factor in the association. Indeed, previous studies which have
evaluated measurement error in these platforms do support this
position\textsuperscript{44,45}, though making strong conclusions for
this analysis is difficult because measurement error is likely to be
study specific.

In our simulations we focused on the simple case of a single instrument
in a single sample setting with a view to making a fair comparison
between MR and the various mediation-based methods available. However,
MR strategies exist for other scenarios, for example if there are
multiple instruments for the hypothesised exposure then the problems
described here may be circumvented\textsuperscript{16}. Likewise, if
there is at least one instrument for each trait then bi-directional MR
can offer solutions to inferring the causal
direction\textsuperscript{46}. In this work we assumed that pleiotropy
(the influence of the instrument on the outcome through a mechanism
other than the exposure) was not present. Recent method developments in
MR\textsuperscript{24,25} have focused on accounting for the issues that
horizontal pleiotropy can introduce, but how they perform in the
presence of measurement error remains to be explored. An important
advantage that MR confers over most mediation based analysis is that it
can be performed in two samples, which can considerably improve power
and expand the scope of analysis. However, whether there is a
substantive difference in two sample MR versus one sample MR in how
measurement error has an effect is not yet fully understood. We have
also assumed no measurement error in the genetic instrument, which is
not unreasonable given the strict QC protocols that ensure high quality
genotype data is available to most studies. We have restricted the scope
to only exploring non-differential measurement error and avoided the
complications incurred if measurement error in the exposure and outcome
is correlated, however our analysis goes beyond many previous
explorations of measurement error by assessing the impacts of both
imprecision (noise) and linear transformations of the true variable on
causal inference.

Mediation based network approaches, that go beyond analyses of two
variables, are very well established\textsuperscript{33} and have a
number of extensions that make them valuable tools, including for
example network construction. But because they are predicated on the
basic underlying principles of mediation they are liable to suffer from
the same issues of measurement error. Recent advances in MR methodology,
for example applying MR to genetical genomics\textsuperscript{47},
multivariate MR\textsuperscript{48} and mediation through
MR\textsuperscript{49--51} may offer more robust alternatives for these
more complicated problems.

@@

Assuming that either \(x \rightarrow y\) or \(y \rightarrow x\), the
causal direction can be inferred by evaluating which of \(\rho_{g, x}\)
and \(\rho_{g, y}\) is larger in magnitude. The Steiger test is a
hypothesis test that provides a p-value for observing the difference in
these correlations under the null hypothesis that they are equal.

Assuming the causal direction is \(x \to y\), two stage MR is formulated
using the following regression models:

@@

In the presence of measurement error in \(x\) and \(y\), however, the
empirical inference of the causal direction will instead be based on
evaluating \(\rho_{g, x_o} > \rho_{g, y_o}\), which can be simplified:

@@

In order to assess how reliable the inference of the causal direction is
in the presence of measurement imprecision, we can evaluate the range of
potential values of measurement error in \(x\) and \(y\) over which the
empirical difference in \(\rho_{g, x_o}\) and \(\rho_{g, y_o}\) would
return the wrong causal direction.

@@

\hypertarget{ref-Millstein2016}{}
18. Millstein, J., Chen, G. K. \& Breton, C. V. Cit : Hypothesis Testing
Software for Mediation Analysis in Genomic Applications.
\emph{Bioinformatics} \textbf{AOP,} 1--10 (2016).

@@

\hypertarget{ref-Pierce2012}{}
35. Pierce, B. L. \& VanderWeele, T. J. The effect of non-differential
measurement error on bias, precision and power in Mendelian
randomization studies. \emph{International Journal of Epidemiology}
\textbf{41,} 1383--1393 (2012).

@@

\hypertarget{ref-Revelle2015}{}
39. Revelle, W. \emph{psych: Procedures for Psychological, Psychometric,
and Personality Research}. (Northwestern University, 2015). at
\textless{}\url{http://cran.r-project.org/package=psych}\textgreater{}

@@

\hypertarget{ref-RCoreTeam2015}{}
41. R Core Team. \emph{R: A Language and Environment for Statistical
Computing}. (R Foundation for Statistical Computing, 2015). at
\textless{}\url{https://www.r-project.org/}\textgreater{}

\hypertarget{ref-Bird2002}{}
42. Bird, A. DNA methylation patterns and epigenetic memory. \emph{Genes
\& development} \textbf{16,} 6--21 (2002).

\hypertarget{ref-Cole2014}{}
43. Cole, D. A. \& Preacher, K. J. Manifest Variable Path Analysis:
Potentially Serious and Misleading Consequences Due to Uncorrected
Measurement Error. \emph{Psychological Methods} \textbf{19,} 300--315
(2014).

\hypertarget{ref-Bose2014}{}
44. Bose, M. \emph{et al.} Evaluation of microarray-based DNA
methylation measurement using technical replicates: the Atherosclerosis
Risk In Communities (ARIC) Study. \emph{BMC Bioinformatics} \textbf{15,}
312 (2014).

\hypertarget{ref-Bryant2011}{}
45. Bryant, P. A. \emph{et al.} Technical Variability Is Greater than
Biological Variability in a Microarray Experiment but Both Are
Outweighed by Changes Induced by Stimulation. \emph{PLoS ONE}
\textbf{6,} e19556 (2011).

\hypertarget{ref-Richmond2014}{}
46. Richmond, R. C. \emph{et al.} Assessing Causality in the Association
between Child Adiposity and Physical Activity Levels: A Mendelian
Randomization Analysis. \emph{PLoS Medicine} \textbf{11,} e1001618
(2014).

\hypertarget{ref-Relton2012}{}
47. Relton, C. L. \& Davey Smith, G. Two-step epigenetic Mendelian
randomization: a strategy for establishing the causal role of epigenetic
processes in pathways to disease. \emph{International journal of
epidemiology} \textbf{41,} 161--76 (2012).

\hypertarget{ref-Burgess2014a}{}
48. Burgess, S., Freitag, D. F., Khan, H., Gorman, D. N. \& Thompson, S.
G. Using multivariable Mendelian randomization to disentangle the causal
effects of lipid fractions. \emph{PloS one} \textbf{9,} e108891 (2014).

\hypertarget{ref-Varbo2015}{}
49. Varbo, A. \emph{et al.} Remnant cholesterol, low-density lipoprotein
cholesterol, and blood pressure as mediators from obesity to ischemic
heart disease. \emph{Circulation research} \textbf{116,} 665--73 (2015).

\hypertarget{ref-Burgess2015}{}
50. Burgess, S., Daniel, R. M., Butterworth, A. S. \& Thompson, S. G.
Network Mendelian randomization: using genetic variants as instrumental
variables to investigate mediation in causal pathways.
\emph{International journal of epidemiology} \textbf{44,} 484--95
(2015).

\hypertarget{ref-Richmond2016}{}
51. Richmond, R. C., Hemani, G., Tilling, K., Davey Smith, G. \& Relton,
C. L. Challenges and novel approaches for investigating molecular
mediation. \emph{Human molecular genetics} \textbf{25,} R149--R156
(2016).
